Looking at Non-Linear Dimension Reductions as Models in the Data Space

### 2025-06-16 paper.qmd
Moved Five Gaussian clusters code to script

training_data_gau <- read_rds("data/five_gau_clusters/data_five_gau.rds")

data_gau <- training_data_gau |> 
  select(-ID) |>
  mutate(type = "data")

tsne_data_gau <- read_rds("data/five_gau_clusters/tsne_data_five_gau_71.rds")

## Compute hexbin parameters
num_bins_x_gau1 <- 18

algo_obj_gau1 <- fit_highd_model(
  highd_data = training_data_gau, 
  nldr_data = tsne_data_gau, 
  bin1 = num_bins_x_gau1, 
  q = 0.1, 
  benchmark_highdens = 5)

tsne_gau_scaled <- algo_obj_gau1$nldr_obj$scaled_nldr
tr_from_to_df_gau1 <- algo_obj_gau1$trimesh_data
df_bin_centroids_gau1 <- algo_obj_gau1$model_2d
df_bin_gau1 <- algo_obj_gau1$model_highd

trimesh_removed_gau_tsne <- ggplot() + 
  geom_point(
    data = tsne_gau_scaled,
    aes(
      x = tSNE1,
      y = tSNE2
    ),
    colour = clr_choice,
    alpha = 0.1) + 
  geom_segment(
    data = tr_from_to_df_gau1,
    aes(
      x = x_from,
      y = y_from,
      xend = x_to,
      yend = y_to),
    colour = "#000000",
    #alpha = 0.4,
    linewidth = 0.5) +
  interior_annotation("a1", 
                      position = c(0.08, 0.95),
                      cex = 2) +
  theme(
    aspect.ratio = 1
  )

df_b <- df_bin_gau1 |>
  dplyr::filter(hexID %in% df_bin_centroids_gau1$hexID) |>
  dplyr::mutate(type = "model") ## Data with summarized mean

## Reorder the rows of df_b according to the hexID order in df_b_with_center_data
df_b <- df_b[match(df_bin_centroids_gau1$hexID, df_b$hexID),] |>
  dplyr::select(-hexID)

# Apply the scaling
df_model_data <- bind_rows(data_gau, df_b)
scaled_gau <- scale_data_manual(df_model_data, "type") |>
  as_tibble()

scaled_gau_data <- scaled_gau |>
  filter(type == "data") |>
  select(-type)

scaled_gau_data_model <- scaled_gau |>
  filter(type == "model") |>
  select(-type)

@@

```{r}
#| label: langevitour-five-gau-tsne-proj
#| eval: false

df_model_data_n <- bind_rows(df_b, data_gau)

langevitour::langevitour(df_model_data_n[1:(length(df_model_data_n)-1)],
                         lineFrom = tr_from_to_df_gau1$from,
                         lineTo = tr_from_to_df_gau1$to,
                         group = factor(df_model_data_n$type,
                                        c("data", "model")),
                         levelColors = c(clr_choice, "#000000"),
                         pointSize = 0.7)
```

```{r}
#| label: five-gau-tsne-projs
projection <- cbind(
  c(0.10394,0.08379,-0.17868,0.32960),
  c(0.35111,-0.01137,-0.09657,-0.16018))

proj_obj1 <- get_projection(projection = projection, 
               proj_scale = 2.5, 
               highd_data = scaled_gau_data, 
               model_highd = scaled_gau_data_model, 
               trimesh_data = tr_from_to_df_gau1, 
               axis_param = list(limits = 1,
                                 axis_scaled = 2,
                                 axis_pos_x = -0.72, 
                                 axis_pos_y = -0.72, 
                                 threshold = 0.1))

projected_df <- proj_obj1$projected_df
model_df <- proj_obj1$model_df
axes <- proj_obj1$axes
circle <- proj_obj1$circle

five_gau_proj_tsne_model1 <- projected_df |>
  ggplot(
    aes(
      x = proj1, 
      y = proj2)) +
  geom_point(
    #size = 0.5,
    alpha = 0.05,
    colour = clr_choice) +
  geom_segment(
    data = model_df, 
    aes(
      x = proj1_from, 
      y = proj2_from, 
      xend = proj1_to, 
      yend = proj2_to), 
    colour = "#000000",
    alpha = 0.4,
    linewidth = 0.8) +
  geom_segment(
    data=axes, 
    aes(x=x1, y=y1, xend=x2, yend=y2), 
    colour="grey70") +
  geom_text(
    data=axes, 
    aes(x=x2, y=y2),
    label=rownames(axes), 
    colour="grey50",
    size = 5) +
  geom_path(
    data=circle, 
    aes(x=c1, y=c2), colour="grey70") +
  coord_fixed() +
  xlim(c(-0.89, 0.89)) +
  ylim(c(-0.89, 0.89)) +
  interior_annotation("a2", 
                      position = c(0.08, 0.9),
                      cex = 2)

## Second projection
projection <- cbind(
  c(0.30943,-0.08452,-0.23005,-0.05284),
  c(0.22463,-0.06682,0.32106,0.02452))

proj_obj2 <- get_projection(projection = projection, 
                            proj_scale = 3, 
                            highd_data = scaled_gau_data, 
                            model_highd = scaled_gau_data_model, 
                            trimesh_data = tr_from_to_df_gau1, 
                            axis_param = list(limits = 1.5,
                                              axis_scaled = 2,
                                              axis_pos_x = -1, 
                                              axis_pos_y = -1, 
                                              threshold = 0.05))

@@

umap_data_gau <- read_rds("data/five_gau_clusters/umap_data_five_gau.rds")

## Compute hexbin parameters
num_bins_x_gau1 <- 63

algo_obj_gau1 <- fit_highd_model(
  highd_data = training_data_gau, 
  nldr_data = umap_data_gau, 
  bin1 = num_bins_x_gau1, 
  q = 0.1, 
  benchmark_highdens = 5)

umap_gau_scaled <- algo_obj_gau1$nldr_obj$scaled_nldr
tr_from_to_df_gau1 <- algo_obj_gau1$trimesh_data
df_bin_centroids_gau1 <- algo_obj_gau1$model_2d
df_bin_gau1 <- algo_obj_gau1$model_highd

trimesh_removed_gau_umap <- ggplot() + 
  geom_point(
    data = umap_gau_scaled,
    aes(
      x = UMAP1,
      y = UMAP2
    ),
    colour = clr_choice,
    alpha = 0.05) + 
  geom_segment(
    data = tr_from_to_df_gau1,
    aes(
      x = x_from,
      y = y_from,
      xend = x_to,
      yend = y_to),
    colour = "#000000",
    #alpha = 0.4,
    linewidth = 0.5) +
  interior_annotation("b1", 
                      position = c(0.95, 0.95),
                      cex = 2) +
  theme(
    aspect.ratio = 1
  )

df_b <- df_bin_gau1 |>
  dplyr::filter(hexID %in% df_bin_centroids_gau1$hexID) |>
  dplyr::mutate(type = "model") ## Data with summarized mean

## Reorder the rows of df_b according to the hexID order in df_b_with_center_data
df_b <- df_b[match(df_bin_centroids_gau1$hexID, df_b$hexID),] |>
  dplyr::select(-hexID)

# Apply the scaling
df_model_data <- bind_rows(data_gau, df_b)
scaled_gau <- scale_data_manual(df_model_data, "type") |>
  as_tibble()

scaled_gau_data <- scaled_gau |>
  filter(type == "data") |>
  select(-type)

scaled_gau_data_model <- scaled_gau |>
  filter(type == "model") |>
  select(-type)

```{r}
#| label: langevitour-five-gau-umap-proj
#| eval: false

df_model_data_n <- bind_rows(df_b, data_gau)

langevitour::langevitour(df_model_data_n[1:(length(df_model_data_n)-1)],
                         lineFrom = tr_from_to_df_gau1$from,
                         lineTo = tr_from_to_df_gau1$to,
                         group = factor(df_model_data_n$type,
                                        c("data", "model")),
                         levelColors = c(clr_choice, "#000000"), pointSize = 0.7)
```

```{r}
#| label: five-gau-umap-projs
projection <- cbind(
  c(0.10394,0.08379,-0.17868,0.32960),
  c(0.35111,-0.01137,-0.09657,-0.16018))

proj_obj1 <- get_projection(projection = projection, 
               proj_scale = 2.5, 
               highd_data = scaled_gau_data, 
               model_highd = scaled_gau_data_model, 
               trimesh_data = tr_from_to_df_gau1, 
               axis_param = list(limits = 1,
                                 axis_scaled = 2,
                                 axis_pos_x = -0.72, 
                                 axis_pos_y = -0.72, 
                                 threshold = 0.1))

projected_df <- proj_obj1$projected_df
model_df <- proj_obj1$model_df
axes <- proj_obj1$axes
circle <- proj_obj1$circle

five_gau_proj_umap_model1 <- projected_df |>
  ggplot(
    aes(
      x = proj1, 
      y = proj2)) +
  geom_point(
    alpha = 0.05,
    colour = clr_choice) +
  geom_segment(
    data = model_df, 
    aes(
      x = proj1_from, 
      y = proj2_from, 
      xend = proj1_to, 
      yend = proj2_to), 
    colour = "#000000",
    alpha = 0.4,
    linewidth = 0.8) +
  geom_segment(
    data=axes, 
    aes(x=x1, y=y1, xend=x2, yend=y2), 
    colour="grey70") +
  geom_text(
    data=axes, 
    aes(x=x2, y=y2, label=rownames(axes)), 
    colour="grey50",
    size = 5) +
  geom_path(
    data=circle, 
    aes(x=c1, y=c2), colour="grey70") +
  coord_fixed() +
  xlim(c(-0.89, 0.89)) +
  ylim(c(-0.89, 0.89)) +
  interior_annotation("b2", 
                      position = c(0.08, 0.9),
                      cex = 2)

## Second projection
projection <- cbind(
  c(0.30943,-0.08452,-0.23005,-0.05284),
  c(0.22463,-0.06682,0.32106,0.02452))

proj_obj2 <- get_projection(projection = projection, 
                            proj_scale = 3, 
                            highd_data = scaled_gau_data, 
                            model_highd = scaled_gau_data_model, 
                            trimesh_data = tr_from_to_df_gau1, 
                            axis_param = list(limits = 1.5,
                                              axis_scaled = 2,
                                              axis_pos_x = -1, 
                                              axis_pos_y = -1, 
                                              threshold = 0.05))

@@

pacmap_data_gau <- read_rds("data/five_gau_clusters/pacmap_data_five_gau.rds")

## Compute hexbin parameters
num_bins_x_gau1 <- 28

algo_obj_gau1 <- fit_highd_model(
  highd_data = training_data_gau, 
  nldr_data = pacmap_data_gau, 
  bin1 = num_bins_x_gau1, 
  q = 0.1, 
  benchmark_highdens = 5)

pacmap_gau_scaled <- algo_obj_gau1$nldr_obj$scaled_nldr
tr_from_to_df_gau1 <- algo_obj_gau1$trimesh_data
df_bin_centroids_gau1 <- algo_obj_gau1$model_2d
df_bin_gau1 <- algo_obj_gau1$model_highd

trimesh_removed_gau_pacmap <- ggplot() + 
  geom_point(
    data = pacmap_gau_scaled,
    aes(
      x = PaCMAP1,
      y = PaCMAP2),
      colour = clr_choice,
    alpha = 0.1) +
  geom_segment(
    data = tr_from_to_df_gau1,
    aes(
      x = x_from,
      y = y_from,
      xend = x_to,
      yend = y_to),
    colour = "#000000",
    #alpha = 0.4,
    linewidth = 0.5) +
  interior_annotation("c1", 
                      position = c(0.08, 0.95),
                      cex = 2) +
  theme(
    aspect.ratio = 1
  )

df_b <- df_bin_gau1 |>
  dplyr::filter(hexID %in% df_bin_centroids_gau1$hexID) |>
  dplyr::mutate(type = "model") ## Data with summarized mean

## Reorder the rows of df_b according to the hexID order in df_b_with_center_data
df_b <- df_b[match(df_bin_centroids_gau1$hexID, df_b$hexID),] |>
  dplyr::select(-hexID)

# Apply the scaling
df_model_data <- bind_rows(data_gau, df_b)
scaled_gau <- scale_data_manual(df_model_data, "type") |>
  as_tibble()

scaled_gau_data <- scaled_gau |>
  filter(type == "data") |>
  select(-type)

scaled_gau_data_model <- scaled_gau |>
  filter(type == "model") |>
  select(-type)
```

```{r}
#| label: langevitour-five-gau-pacmap-proj
#| eval: false

df_model_data_n <- bind_rows(df_b, data_gau)

langevitour::langevitour(df_model_data_n[1:(length(df_model_data_n)-1)],
                         lineFrom = tr_from_to_df_gau1$from,
                         lineTo = tr_from_to_df_gau1$to,
                         group = factor(df_model_data_n$type,
                                        c("data", "model")),
                         levelColors = c(clr_choice, "#000000"), pointSize = 0.7)
```

```{r}
#| label: five-gau-pacmap-projs
## First projection
projection <- cbind(
  c(0.10394,0.08379,-0.17868,0.32960),
  c(0.35111,-0.01137,-0.09657,-0.16018))

proj_obj1 <- get_projection(projection = projection, 
               proj_scale = 2.5, 
               highd_data = scaled_gau_data, 
               model_highd = scaled_gau_data_model, 
               trimesh_data = tr_from_to_df_gau1, 
               axis_param = list(limits = 1,
                                 axis_scaled = 2,
                                 axis_pos_x = -0.72, 
                                 axis_pos_y = -0.72, 
                                 threshold = 0.1))

projected_df <- proj_obj1$projected_df
model_df <- proj_obj1$model_df
axes <- proj_obj1$axes
circle <- proj_obj1$circle

five_gau_proj_pacmap_model1 <- projected_df |>
  ggplot(
    aes(
      x = proj1,
      y = proj2)) +
  geom_point(
    alpha = 0.05,
    colour = clr_choice) +
  geom_segment(
    data = model_df,
    aes(
      x = proj1_from,
      y = proj2_from,
      xend = proj1_to,
      yend = proj2_to),
    colour = "#000000",
    alpha = 0.4,
    linewidth = 0.8) +
  geom_segment(
    data=axes,
    aes(x=x1, y=y1, xend=x2, yend=y2),
    colour="grey70") +
  geom_text(
    data=axes,
    aes(x=x2, y=y2, label=rownames(axes)),
    colour="grey50",
    size = 5) +
  geom_path(
    data=circle,
    aes(x=c1, y=c2), colour="grey70") +
  coord_fixed() +
  xlim(c(-0.89, 0.89)) +
  ylim(c(-0.89, 0.89)) +
  interior_annotation("c2",
                      position = c(0.08, 0.9),
                      cex = 2)

## Second projection
projection <- cbind(
  c(0.30943,-0.08452,-0.23005,-0.05284),
  c(0.22463,-0.06682,0.32106,0.02452))

proj_obj2 <- get_projection(projection = projection, 
                            proj_scale = 3, 
                            highd_data = scaled_gau_data, 
                            model_highd = scaled_gau_data_model, 
                            trimesh_data = tr_from_to_df_gau1, 
                            axis_param = list(limits = 1.5,
                                              axis_scaled = 2,
                                              axis_pos_x = -1, 
                                              axis_pos_y = -1, 
                                              threshold = 0.05))

>>>>

tsne_gau_scaled <- read_rds("data/five_gau_clusters/tsne_gau_scaled.rds")
tr_from_to_df_gau1 <- read_rds("data/five_gau_clusters/tr_from_to_df_gau1.rds")

trimesh_removed_gau_tsne <- ggplot() + 
  geom_point(
    data = tsne_gau_scaled,
    aes(
      x = tSNE1,
      y = tSNE2
    ),
    colour = clr_choice,
    alpha = 0.1) + 
  geom_segment(
    data = tr_from_to_df_gau1,
    aes(
      x = x_from,
      y = y_from,
      xend = x_to,
      yend = y_to),
    colour = "#000000",
    linewidth = 0.5) +
  interior_annotation("a1", 
                      position = c(0.08, 0.95),
                      cex = 2) +
  theme(
    aspect.ratio = 1
  )

@@

```{r}
#| label: five-gau-tsne-projs
## First projection
proj_obj1 <- read_rds("data/five_gau_clusters/tsne_gau_proj_obj1.rds")

five_gau_proj_tsne_model1 <- plot_proj(
  proj_obj = proj_obj1,
  point_param = c(1.5, 0.05, clr_choice),
  line_param = c(0.8, 0.4, "#000000"),
  plot_limits = c(-0.89, 0.89),
  axis_text_size = 5,
  is_category = FALSE
) +
  interior_annotation("a2", 
                      position = c(0.08, 0.9),
                      cex = 2)

## Second projection

proj_obj2 <- read_rds("data/five_gau_clusters/tsne_gau_proj_obj2.rds")

@@

umap_gau_scaled <- read_rds("data/five_gau_clusters/umap_gau_scaled.rds")
tr_from_to_df_gau1_umap <- read_rds("data/five_gau_clusters/tr_from_to_df_gau1_umap.rds")

trimesh_removed_gau_umap <- ggplot() + 
  geom_point(
    data = umap_gau_scaled,
    aes(
      x = UMAP1,
      y = UMAP2
    ),
    colour = clr_choice,
    alpha = 0.05) + 
  geom_segment(
    data = tr_from_to_df_gau1_umap,
    aes(
      x = x_from,
      y = y_from,
      xend = x_to,
      yend = y_to),
    colour = "#000000",
    #alpha = 0.4,
    linewidth = 0.5) +
  interior_annotation("b1", 
                      position = c(0.95, 0.95),
                      cex = 2) +
  theme(
    aspect.ratio = 1
  )
```

```{r}
#| label: five-gau-umap-projs

## First projection
proj_obj1 <- read_rds("data/five_gau_clusters/umap_gau_proj_obj1.rds")

five_gau_proj_umap_model1 <- plot_proj(
  proj_obj = proj_obj1,
  point_param = c(1.5, 0.05, clr_choice),
  line_param = c(0.8, 0.4, "#000000"),
  plot_limits = c(-0.89, 0.89),
  axis_text_size = 5,
  is_category = FALSE
) +
  interior_annotation("b2", 
                      position = c(0.08, 0.9),
                      cex = 2)

## Second projection
proj_obj2 <- read_rds("data/five_gau_clusters/umap_gau_proj_obj2.rds")

@@

pacmap_gau_scaled <- read_rds("data/five_gau_clusters/pacmap_gau_scaled.rds")
tr_from_to_df_gau1_pacmap <- read_rds("data/five_gau_clusters/tr_from_to_df_gau1_pacmap.rds")

trimesh_removed_gau_pacmap <- ggplot() + 
  geom_point(
    data = pacmap_gau_scaled,
    aes(
      x = PaCMAP1,
      y = PaCMAP2),
      colour = clr_choice,
    alpha = 0.1) +
  geom_segment(
    data = tr_from_to_df_gau1_pacmap,
    aes(
      x = x_from,
      y = y_from,
      xend = x_to,
      yend = y_to),
    colour = "#000000",
    #alpha = 0.4,
    linewidth = 0.5) +
  interior_annotation("c1", 
                      position = c(0.08, 0.95),
                      cex = 2) +
  theme(
    aspect.ratio = 1
  )
```

```{r}
#| label: five-gau-pacmap-projs
## First projection
proj_obj1 <- read_rds("data/five_gau_clusters/pacmap_gau_proj_obj1.rds")

five_gau_proj_pacmap_model1 <- plot_proj(
  proj_obj = proj_obj1,
  point_param = c(1.5, 0.05, clr_choice),
  line_param = c(0.8, 0.4, "#000000"),
  plot_limits = c(-0.89, 0.89),
  axis_text_size = 5,
  is_category = FALSE
) +
  interior_annotation("c2",
                      position = c(0.08, 0.9),
                      cex = 2)

## Second projection
proj_obj2 <- read_rds("data/five_gau_clusters/pacmap_gau_proj_obj2.rds")
